paper figures

### 2025-07-31 paper_figures.Rmd
new panels for revision

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
#library(DescTools)
library(lme4)
```

@@

columns <- list('Diversity'=c('nclusters_final_downsampled',
                              'nclusters_eff_downsampled',
                              'mean_cluster_size_downsampled',
                              'std_cluster_size_downsampled',
                              'cluster_gini_downsampled',
                              'cluster_entropy_downsampled',
                              'cluster_inverse_simpson_downsampled',
                              'top_clone_occupancy_downsampled',
                              'big_clones_occupancy_downsampled'),

@@

                             'frac_clusters_SA1','frac_clusters_SA2','frac_clusters_SE',
                             'frac_clusters_SG1','frac_clusters_SG2','frac_clusters_SG3',
                             'frac_clusters_SG4','frac_clusters_SM',
                             'frac_clusters_Sa','frac_clusters_Sg3',
                             'frac_clusters_Sg2b','frac_clusters_Sg2c',
                             'frac_clusters_Sm','frac_clusters_Sg1'),
                'Structural'=c('insert_frequency',
                               'frac_breaks_inversions','frac_breaks_duplications',
                               'median_duplication_size','median_inversion_size'),
                'Context'=c('homology_fw','homology_rv',

@@

                            'n_homology_switch', 
                            'n_untemplated_switch',
                            'frac_blunt'),
                'Breaks'=c('mean_length',
                           'std_length',
                           'mean_GC',
                           'std_GC',
                           'mean_length_SA',
                           'mean_length_SG',
                           'std_length_SA',
                           'std_length_SG',
                           'frac_breaks_single',
                           'frac_breaks_sequential',
                           'frac_breaks_intra',
                           'mean_intra_break_size',
                           'std_intra_break_size',
                           'spread_SM','spread_SA','spread_SG',
                           'spread_SA1','spread_SA2',
                           'spread_SG1','spread_SG2',
                           'spread_SG2B','spread_SG2C',
                           'spread_SG3','spread_SG4',
                           'frac_breaks_SM_SM','frac_breaks_SM_SG',
                           'frac_breaks_SG_SG','frac_breaks_SM_SA',
                           'frac_breaks_SG_SA','frac_breaks_SA_SA'),
                'Context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',
                                    'homology_fw_SM_SG','homology_rv_SM_SG',

@@

                                    'n_homology_switch_SA', 
                                    'n_homology_switch_SG', 
                                    'n_untemplated_switch_SA',
                                    'n_untemplated_switch_SG',
                                    'frac_blunt_SA',
                                    'frac_blunt_SG'),
                'Diversity_raw'=c('nclusters_final',
                                  'nclusters_eff',
                                  'mean_cluster_size',
                                  'std_cluster_size',
                                  'top_clone_occupancy',
                                  'big_clones_occupancy'),
                'Variants'=c('frac_variants_germline',

@@

                                levels=c('Library',
                                         'Isotypes',

@@

cat_cols <- c('Structural'='#FD9E40',
              'Context'='#FED6AC',
              'Context_isotype'='#FED6AC',
              'Breaks'='#1E8AC4',
              'Isotypes'='#AEF58E',
              'Diversity_raw'='#3A7A2C',
              'Diversity'='#3A7A2C',
              'Variants'='lightpink3',
              'Library'='gray40')

@@

tissue_colors <- c('Splenocytes'='orange2','Lymph Node'='orangered3','CH12'='forestgreen')
var_colors <- c('Batch'='palegreen2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
                'Sex'='steelblue2','Age'='darkseagreen4')

@@

group_colors <- c('SmB+CD21n'='gray20',
                  'SmB-CD21lo'='orange3',
                  'SmB-CD21n'='lightseagreen')

col_cols <- columns %>% 
  dplyr::mutate(color=cat_cols[as.character(category)]) %>%
  dplyr::mutate(col=gsub('_downsampled|_switch','',col)) %>%
  dplyr::pull(color, col)

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=nclusters_final, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
```

@@

ggplot(cellswitch.data %>%
         dplyr::mutate(Donor=plyr::revalue(Donor,
                                           c('21084'='A',
                                             '21085'='B',
                                             '21086'='C'))),
       aes(x=ncells, y=nclusters_final)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(aes(color=Donor),size=.5) + 
  geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(legend.position=c(.8,.25),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Input cells', y='# Filtered clusters', color='Donor')
```

@@

```{r Fig1F, fig.width=2.4 * fig_scale, fig.height=1.4 * fig_scale}
# bulk VDJ results for Fig. 1
VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)

vdj_stats <- VDJ_results %>%
  dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
  dplyr::group_by(Donor,ncells,Sample) %>%
  dplyr::summarise(nclusters_final=n_distinct(clone),
                   alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
  dplyr::mutate(method='VDJ')

swibrid_stats <- cellswitch.data %>%
  dplyr::select(Donor,ncells,
                nclusters_final,
                alpha_ratio_clusters) %>%
  dplyr::filter(ncells %in% c(50000,100000)) %>%
  dplyr::mutate(method='SWIBRID')

@@

vdj_df <- rbind(swibrid_stats[,vdj_cols],vdj_stats[,vdj_cols]) %>%
  gather(metric,value, c(nclusters_final,alpha_ratio_clusters)) %>%
  dplyr::mutate(ncells=factor(as.character(as.integer(ncells)), levels=c('50000','100000')))

# vdj_df %>% 
#   dplyr::filter(metric=='nclusters_final') %>%
#   ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
#   geom_boxplot(outlier.shape=NA, lwd=.5) +
#   geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
#   facet_wrap(~method, ncol=2, scales='free_y') + 
#   my_theme() +
#   scale_color_brewer(palette='Set1') +
#   theme(strip.clip='off',
#         legend.position='none',
#         legend.background =element_rect(fill=NA)) +
#   labs(y='# Clones',x='# Cells / 1000',color='')

@@

@@

cols.here <- c('cluster_entropy_downsampled',
               'cluster_gini_downsampled',
               'mean_cluster_size_downsampled',
               'std_cluster_size_downsampled',
               'top_clone_occupancy_downsampled',
               'alpha_ratio_clusters',
               'frac_clusters_SA1',
               'frac_clusters_SA2',
               'frac_clusters_SG1',
               'frac_clusters_SG2',
               'insert_frequency',
               'frac_breaks_inversions',
               'frac_breaks_duplications',
               'median_duplication_size',
               'median_inversion_size',
               'homology_fw',
               'donor_score_WGCW',
               'acceptor_complexity',
               'n_homology_switch',
               'n_untemplated_switch',
               'frac_breaks_single',
               'frac_breaks_sequential',
               'frac_breaks_intra',
               'spread_SM',
               'spread_SA')

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','unexplained'))),
                category=factor(category, levels=c('Structural','Context','Breaks','Diversity','Isotypes'))) %>%
  dplyr::filter(col %in% cols.here) %>%
  dplyr::mutate(col=factor(gsub('_downsampled|_switch','',col), levels=gsub('_downsampled|_switch','',cols.here))) %>%
  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
  geom_col(width=.8) +
  my_theme() +
  facet_wrap(~category,scales='free_y', ncol=5) +
  theme(strip.clip='off',
        legend.position='bottom',
        legend.box='horizontal') +
  guides(fill=guide_legend(ncol=6)) +
  coord_flip() +
  scale_y_continuous(breaks=c(0,50,100)) +
  scale_fill_manual(values=var_colors,
                    breaks=c('Donor','Sex','Age','Batch','Sequencing','unexplained')) +
  labs(x='', y='% Explained variance', fill='')
```

```{r Fig2D, fig.width=1.75 * fig_scale,fig.height=1.7 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                Donor >= donor_cutoff, 
                Sequencing < sequencing_cutoff, 
                Batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

@@

```{r Fig2E, fig.width=.9 * fig_scale, fig.height=1.7 * fig_scale}
age.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
  dplyr::slice_max(Age, n=1) %>%
  dplyr::pull(col)
age.cols <- 'top_clone_occupancy_downsampled'

sex.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
  dplyr::slice_max(Sex, n=1) %>%
  dplyr::pull(col)
sex.cols <- 'spread_SG1'

HD.data %>%
  dplyr::filter(feature %in% sex.cols, Cohort=='C1') %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                               levels=gsub('_downsampled|_switch','',sex.cols))) %>% 
  ggplot(aes(x=str_to_title(Sex), y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('Male','Female')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
  my_theme() + 
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y='')

HD.data %>%
  dplyr::filter(feature %in% age.cols, Cohort=='C1') %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature),
                               levels=gsub('_downsampled|_switch','',age.cols))) %>% 
  ggplot(aes(x=Age, y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('<50','>=50')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  facet_wrap(~feature, scales='free', ncol=8, switch='y') + 
  my_theme() +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y='')
```

@@

R <- cor(FB.X[,selected.cols], use='na.or.complete')
io <- gsub('_downsampled|_switch','',rownames(R)[order.hclust(hclust(dist(R)))])

tmp <- as.data.frame(R) %>% 
  tibble::rownames_to_column('var2') %>%
  gather(var1, R, -var2) %>%
  dplyr::left_join(columns, by=c('var1'='col')) %>%
  dplyr::mutate(cat1=category) %>%
  dplyr::select(-category) %>%
  dplyr::left_join(columns, by=c('var2'='col')) %>%
  dplyr::mutate(cat2=category) %>%
  dplyr::select(-category)  %>%
  dplyr::mutate(var1=factor(gsub('_downsampled|_switch','',var1), levels=io),
                var2=factor(gsub('_downsampled|_switch','',var2), levels=io),
                i1=as.vector(sapply(var1, function(x) which(io==x))),
                i2=as.vector(sapply(var2, function(x) which(io==x))))

@@

ggplot(tmp %>%
  dplyr::filter(i1 <= i2),
  aes(x=var1,y=var2,fill=R)) + 
  geom_tile() +
  scale_fill_gradient2(low='blue3',high='red3',mid='gray',limits=c(-1,1)) + 
  my_theme() +
  scale_x_discrete(position = "top") +
  coord_cartesian(clip='off') +
  theme(axis.text.x.top=element_text(angle=90, hjust=0., vjust=.5), #, colour=col_cols[io]),
        legend.position=c(-.1,1.2)) +
  labs(x='',y='') +
  annotation_custom(grob=anno_grob_y, xmin = 0, xmax = 1, ymin = length(io) + 10, ymax = length(io) + 28) +
  annotation_custom(grob=anno_grob_x, xmin = -9, xmax = 8, ymin = 0, ymax = 1)

@@

```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
library(gtools)
cols.here.human <- c('insert_frequency',
                     'n_untemplated_switch',
                     'frac_breaks_sequential',
                     'frac_breaks_intra',
                     'cluster_gini_downsampled',
                     'frac_clusters_SG1')

cohort.data <- HD.data %>%
  dplyr::filter(Sequencing=='minION') %>% 
  dplyr::filter(feature %in% cols.here.human) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=factor(gsub('_downsampled|_switch','',feature), 
                               levels=gsub('_downsampled|_switch','',cols.here.human))) %>%
  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))

@@

```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
library(gtools)
mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')
vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')

cols.here.mouse <- c('frac_breaks_inversions',
                     'frac_breaks_intra',
                     'mean_cluster_size_downsampled')

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
library(gtools)
library(lme4)
library(dendextend)
```

@@

columns <- list('Diversity'=c('clusters',
                              'clusters_eff',
                              'cluster_size_mean',
                              'cluster_size_std',
                              'occupancy_top_cluster',
                              'occupancy_big_clusters'),

@@

                             'pct_clusters_SA1','pct_clusters_SA2','pct_clusters_SE',
                             'pct_clusters_SG1','pct_clusters_SG2','pct_clusters_SG3',
                             'pct_clusters_SG4','pct_clusters_SM',
                             'pct_clusters_Sa','pct_clusters_Sg3',
                             'pct_clusters_Sg2b','pct_clusters_Sg2c',
                             'pct_clusters_Sm','pct_clusters_Sg1'),
                'Structural'=c('pct_templated_inserts',
                               'pct_inversions','pct_duplications',
                               'duplication_size','inversion_size'),
                'Context'=c('homology_score_fw','homology_score_rv',

@@

                            'homology', 
                            'untemplated_inserts',
                            'pct_blunt'),
                'Breaks'=c('read_length_mean',
                           'read_length_std',
                           'GC_content_mean',
                           'GC_content_std',
                           'read_length_SA_mean',
                           'read_length_SG_mean',
                           'read_length_SA_std',
                           'read_length_SG_std',
                           'pct_direct_switch',
                           'pct_sequential_switch',
                           'pct_intraswitch_deletion',
                           'intraswitch_size_mean',
                           'intraswitch_size_std',
                           'break_dispersion_SM','break_dispersion_SA','break_dispersion_SG',
                           'break_dispersion_SA1','break_dispersion_SA2',
                           'break_dispersion_SG1','break_dispersion_SG2',
                           'break_dispersion_SG2B','break_dispersion_SG2C',
                           'break_dispersion_SG3','break_dispersion_SG4',
                           'pct_breaks_SM_SM','pct_breaks_SM_SG',
                           'pct_breaks_SG_SG','pct_breaks_SM_SA',
                           'pct_breaks_SG_SA','pct_breaks_SA_SA'),
                'Context_isotype'=c('homology_score_fw_SM_SA','homology_score_rv_SM_SA',
                                    'homology_score_fw_SM_SG','homology_score_rv_SM_SG',

@@

                                    'homology_SA', 
                                    'homology_SG', 
                                    'untemplated_inserts_SA',
                                    'untemplated_inserts_SG',
                                    'pct_blunt_SA',
                                    'pct_blunt_SG'),
                'Diversity_raw'=c('clusters_raw',
                                  'clusters_eff_raw',
                                  'cluster_size_mean_raw',
                                  'cluster_size_std_raw',
                                  'cluster_gini_raw',
                                  'cluster_entropy_raw',
                                  'cluster_inverse_simpson_raw',
                                  'occupancy_top_cluster_raw',
                                  'occupancy_big_clusters_raw'),
                'Variants'=c('somatic_variants',

@@

                                levels=c('Isotypes',

@@

cat_cols <- c('Structural'='#FD9E40',
              'Context'='#FED6AC',
              'Context_isotype'='#FED6AC',
              'Breaks'='#1E8AC4',
              'Isotypes'='#AEF58E',
              'Diversity_raw'='#3A7A2C',
              'Diversity'='#3A7A2C',
              'Variants'='lightpink3',
              'Library'='gray40')

@@

tissue_colors <- c('Splenocytes'='orange2','Lymph Node'='orangered3','CH12'='forestgreen')
var_colors <- c('Batch'='lavenderblush2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
                'Sex'='steelblue2','Age'='lightgoldenrod1','Treatment'='saddlebrown','IV'='navy')

@@

col_cols <- columns %>% 
  dplyr::mutate(color=cat_cols[as.character(category)]) %>%
  dplyr::pull(color, col)

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=clusters_raw, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
```

@@

ggplot(cellswitch.data %>%
         dplyr::mutate(Donor=plyr::revalue(Donor,
                                           c('21084'='A',
                                             '21085'='B',
                                             '21086'='C'))),
       aes(x=ncells, y=clusters_raw)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(aes(color=Donor),size=.5) + 
  geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(legend.position=c(.8,.25),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Input cells', y='# Filtered clusters', color='Donor')
```

check reproducibility of clones in the cell switch experiment

```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
meta_clustering <- read.csv('data/cellswitch_meta_clustering.csv', header=1, row.names=1)

ggplot(meta_clustering %>% 
         dplyr::mutate(ncells=factor(ncells, levels=c('50k','100k')),
                       Donor=ifelse(Donor=='mixed','mixed',paste0(Donor,'-',Donor)),
                       mixing=gsub('-donor','',mixing)), 
       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
  geom_jitter(aes(color=Donor, group=mixing), position=position_jitterdodge(dodge.width = .8, jitter.width=.05), size=1) +
  my_theme() +
  #theme(legend.position='bottom') +
  #coord_cartesian(ylim=c(15,40)) +
  scale_fill_manual(values=c('gray','white'), guide=FALSE) +
  scale_color_manual(values=setNames(c(brewer.pal(3,'Set1'),'gray20'),c('A-A','B-B','C-C','mixed')),na.value='gray20') +
  labs(x='# cells',y='% shared clusters in pair', fill='pair', color='Donors')
```

@@

```{r Fig1G, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# bulk VDJ results for Fig. 1
VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)

get_gini <- function(x) {
  # calculate Gini coefficient of array x; see https://stackoverflow.com/questions/48999542/more-efficient-weighted-gini-coefficient-in-python
  sorted_x <- sort(x)
  n <- length(x)
  cumx <- cumsum(sorted_x)
  (n + 1 - 2 * sum(cumx) / cumx[n]) / n
}

get_inverse_simpson <- function(x) {
  rel_x <- x/sum(x)
  sum(1 / rel_x**2)
}

vdj_stats <- VDJ_results %>%
  dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
  dplyr::group_by(Donor,ncells,Sample) %>%
  dplyr::summarise(clusters_raw=n_distinct(clone),
                   cluster_gini_raw=get_gini(uniqueMoleculeCount),
                   cluster_inverse_simpson_raw=get_inverse_simpson(uniqueMoleculeCount),
                   alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
  dplyr::mutate(method='VDJ')

swibrid_stats <- cellswitch.data %>%
  dplyr::select(Donor,ncells,
                clusters_raw,
                cluster_gini_raw,
                cluster_inverse_simpson_raw,
                alpha_ratio_clusters) %>%
  dplyr::filter(ncells %in% c(50000,100000)) %>%
  dplyr::mutate(method='SWIBRID')

@@

vdj_df <- rbind(swibrid_stats[,vdj_cols],vdj_stats[,vdj_cols]) %>%
  gather(metric,value, c(clusters_raw,alpha_ratio_clusters,cluster_gini_raw,cluster_inverse_simpson_raw)) %>%
  dplyr::mutate(ncells=factor(as.character(as.integer(ncells)), levels=c('50000','100000')))

@@

```{r FigS1X1, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
vdj_df %>% 
  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
                Donor=gsub('Donor ','',Donor)) %>%
  dplyr::filter(metric=='cluster_gini_raw') %>%
  dplyr::group_by(ncells, Donor, method) %>%
  dplyr::summarise(mean_gini=mean(value),
                   se_gini=sd(value)/sqrt(length(value))) %>%
  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_gini','se_gini')) %>%
  ggplot(aes(x=mean_gini_VDJ, y=mean_gini_SWIBRID,
             xmin=mean_gini_VDJ-se_gini_VDJ,
             xmax=mean_gini_VDJ+se_gini_VDJ,
             ymin=mean_gini_SWIBRID-se_gini_SWIBRID,
             ymax=mean_gini_SWIBRID+se_gini_SWIBRID)) +
  geom_point(aes(color=Donor,shape=ncells),size=1) +
  geom_errorbar(linewidth=.25) +
  geom_errorbarh(linewidth=.25) +
  #scale_x_continuous(breaks=c(.6,.65)) +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.background=element_blank()) +
  labs(y='Gini coefficient SWIBRID',x='Gini coefficient VDJ',shape='# cells')
```

```{r FigS1X2, fig.width=2 * fig_scale, fig.height=1.4 * fig_scale}
library(ggpmisc)
PCR.data <- read.csv('data/PCR_data.csv',header=1,row.names=1)
ggplot(PCR.data, aes(x=as.integer(ncycles), y=log10(reads))) + 
  geom_point(aes(color=donor)) + 
  geom_smooth(method='lm', color='black', linewidth=.5) +
  geom_abline(slope=log10(2), intercept=-1.5, linewidth=.5, linetype='dashed') +
  my_theme() + 
  scale_color_brewer(palette='Set2') +
  scale_y_continuous(breaks = log10(c(1, 10, 100, 1000,10000, 100000)),
                     label = c(1, 10, 100, 1000, 10000, 100000)) +
  labs(x='# of cycles',y='# of reads')
```

@@

cols.here <- c('cluster_entropy',
               'cluster_gini',
               'cluster_size_mean',
               'cluster_size_std',
               'occupancy_top_cluster',
               'alpha_ratio_clusters',
               'pct_clusters_SA1',
               'pct_clusters_SA2',
               'pct_clusters_SG1',
               'pct_clusters_SG2',
               'pct_templated_inserts',
               'pct_inversions',
               'pct_duplications',
               'duplication_size',
               'inversion_size',
               'homology_score_fw',
               'donor_score_WGCW',
               'acceptor_complexity',
               'homology',
               'untemplated_inserts',
               'pct_direct_switch',
               'pct_sequential_switch',
               'pct_intraswitch_deletion',
               'break_dispersion_SM',
               'break_dispersion_SA')

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','unexplained'))),
                category=factor(category, levels=c('Structural','Context','Breaks','Diversity','Isotypes'))) %>%
  dplyr::filter(col %in% cols.here) %>%
  dplyr::mutate(col=factor(col, levels=cols.here)) %>%
  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
  geom_col(width=.8) +
  my_theme() +
  facet_wrap(~category,scales='free_y', ncol=5) +
  theme(strip.clip='off',
        legend.position='bottom',
        legend.box='horizontal') +
  guides(fill=guide_legend(ncol=6)) +
  coord_flip() +
  scale_y_continuous(breaks=c(0,50,100)) +
  scale_fill_manual(values=var_colors,
                    breaks=c('Donor','Sex','Age','Batch','Sequencing','unexplained')) +
  labs(x='', y='% Explained variance', fill='')
```

```{r Fig2D, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                Donor >= donor_cutoff, 
                Sequencing < sequencing_cutoff, 
                Batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

@@

```{r Fig2E, fig.width=.75 * fig_scale, fig.height=1.5 * fig_scale}
col <- 'break_dispersion_SG1'
HD.data %>%
  dplyr::filter(feature == col, Cohort=='C1') %>%
  ggplot(aes(x=str_to_title(Sex), y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('Male','Female')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  my_theme() + 
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y=col)

col <- 'occupancy_top_cluster'
HD.data %>%
  dplyr::filter(feature==col, Cohort=='C1') %>%
  ggplot(aes(x=Age, y=value)) + 
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_jitter(width=.2, size=.1) + 
  stat_compare_means(comparisons=list(c('<50','>=50')),
                     method='wilcox.test',label = "p.signif", hide.ns=TRUE, size=2.5) +
  my_theme() +
  coord_cartesian(clip='off') +
  theme(legend.position='none',
        axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
  labs(x='',y=col)
```

@@

R <- cor(FB.X[,selected.cols], use='na.or.complete')
io <- rownames(R)[order.hclust(hclust(dist(R)))]

tmp <- as.data.frame(R) %>% 
  tibble::rownames_to_column('var2') %>%
  gather(var1, R, -var2) %>%
  dplyr::left_join(columns, by=c('var1'='col')) %>%
  dplyr::mutate(cat1=category) %>%
  dplyr::select(-category) %>%
  dplyr::left_join(columns, by=c('var2'='col')) %>%
  dplyr::mutate(cat2=category) %>%
  dplyr::select(-category)  %>%
  dplyr::mutate(var1=factor(var1, levels=io),
                var2=factor(var2, levels=io),
                i1=as.vector(sapply(var1, function(x) which(io==x))),
                i2=as.vector(sapply(var2, function(x) which(io==x))))

@@

ggplot(tmp %>%
  dplyr::filter(i1 <= i2),
  aes(x=var1,y=var2,fill=R)) + 
  geom_tile() +
  scale_fill_gradient2(low='blue3',high='red3',mid='gray',limits=c(-1,1)) + 
  my_theme() +
  scale_x_discrete(position = "top") +
  coord_cartesian(clip='off') +
  theme(axis.text.x.top=element_text(angle=90, hjust=0., vjust=.5), #, colour=col_cols[io]),
        legend.position=c(-.1,1.2)) +
  labs(x='',y='') +
  annotation_custom(grob=anno_grob_y, xmin = 0, xmax = 1, ymin = length(io) + 10, ymax = length(io) + 28) +
  annotation_custom(grob=anno_grob_x, xmin = -8, xmax = 9, ymin = 0, ymax = 1)

@@

```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
cols.here.human <- c('pct_templated_inserts',
                     'untemplated_inserts',
                     'pct_sequential_switch',
                     'pct_intraswitch_deletion',
                     'cluster_gini',
                     'pct_clusters_SG1')

cohort.data <- HD.data %>%
  dplyr::filter(Sequencing=='minION') %>% 
  dplyr::filter(feature %in% cols.here.human) %>%
  dplyr::left_join(columns, by=c('feature'='col')) %>%
  dplyr::mutate(feature=factor(feature, levels=cols.here.human)) %>%
  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))

@@

```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')
vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')

cols.here.mouse <- c('pct_inversions',
                     'pct_intraswitch_deletion',
                     'cluster_size_mean')

cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     Cohort=material) %>%
                       dplyr::select(species, Cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     Cohort=Genotype) %>%
                       dplyr::select(species, Cohort, cols.here.mouse)) %>%
  dplyr::mutate(Cohort=factor(Cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, Cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::mutate(variable=factor(variable, levels=cols.here.mouse)) %>%
  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))
